Too Beautiful to be Fake: Attractive Faces are Less Likely to be Judged as Artificially Generated

### 2024-12-06 submission/R2/manuscript.qmd
changes to manuscript

**Advancements in technology have now made it possible to create
near-perfect simulations that are indistinguishable from reality with an
ease, affordability and accessibility that are unprecedented in Human
history**. These artificial, yet realistic constructs permeate all areas
of life through immersive works of fiction, deep fakes (real-like images
and videos generated by deep learning algorithms), virtual and augmented
reality (VR and AR), artificial beings (artificial intelligence "bots"
with or without a physical form), fake news and skewed narratives, of
which ground truth is often hard to access [@nightingale2022]. Such
developments not only carry important consequences for the technological
and entertainment sectors, but also for security and politics - for
instance if used for propaganda and disinformation, recruitment into
malevolent organizations, or religious indoctrination [@pantserev2020].
This issue is central to what has been coined the "post-truth era"
[@lewandowsky2017beyond], in which the distinction (and lack thereof)
between authentic and simulated objects will play a critical role.

@@

**While not all simulations have achieved perfect realism, such as
Computer Generated Images (CGI) in movies or via recent algorithms such
as GANs or diffusion model, which often include distortions or lack
certain key details distinguishing them from real images
[@mcdonnell2010face; @corvi2022detection], it is fair to assume that
these technical limitations will become negligible in the near future.**
This is particularly true in the field of face generation, where
face-generation algorithms are already able to create stimuli that are
virtually indistinguishable from real photos [@nightingale2022;
@tucciarelli2020; @moshel2022]. Such a technological feat, however,
leads to a new question: if real and fake stimuli cannot be
differentiated based on their objective "physical" characteristics, how
can we form judgements regarding their nature?

@@

Literature shows that the context surrounding a stimulus often plays an
important role in the assessment of its reality [a process henceforth
referred to as *simulation monitoring*, @makowski2018cognitive;
@makowski2019phenomenal]. With the extensive search and processing of
cues within ambiguous stimuli being an increasingly complex and
cognitively effortful strategy [@michael2021source;
@susmann2021persuasion], people tend to draw on peripheral contextual
cues (**Figure 1**), such as the source of the stimulus (e.g., which
journal was the information published in), and its credibility,
authority and expertise, to help facilitate their evaluation
[@petty1986elaboration; @susmann2021persuasion; @michael2021source].
However, the atomization and decontextualization of information allowed
by online social media (where text snippets or video excerpts are often
mass-shared with little context) makes this task progressively difficult
[@berghel2018weaponizing; @chen2015news]. Thus, in the absence of clear
contextual information, what drives our beliefs of reality?

@@

Evidence suggests that inter-individual characteristics play a crucial
role in simulation monitoring, with factors such as cognitive style,
prior beliefs, and personality traits [@bryanov2021determinants;
@ecker2022psychological; @sindermann2020short]. For instance,
individuals with stronger analytical reasoning skills have been found to
better discriminate real from fake stimuli [@pehlivanoglu2021role;
@pennycook2019lazy], and prior knowledge or beliefs about the stimulus
influences one's perception of it by biasing the attention deployment
towards information that is in line with one's expectations
[@britt2019reasoned]. Furthermore, dispositional traits, such as high
levels of narcissism and low levels of openness and conscientiousness,
have been associated with greater susceptibility to fake news
[@piksa2022cognitive; @sindermann2020short]. **Interestingly, a recent
review suggested that narcissism was related to a strong self-perceived
ability at detecting lies [@turi2022tangled], which could translate to
participants scoring high on narcissism providing more clear cut and
confidence responses. Conversely, those high in honesty-humility tend to
be more conservative in their judgments to ensure fairness
[@liu2020honesty], likely resulting in lower confidence ratings.**

@@

Beyond stimulus- and individual-related characteristics, evidence
suggests that the interaction between the two (i.e., the subjective
reaction associated with the experience of a given stimulus),
contributes to simulation monitoring decisions. For instance, the
intensity of experienced emotions have been shown to increase one's
sense of presence - the extent to which one feels like "being there", as
if the object of experience was real - when engaged in a fictional movie
or a VR environment [@sanchez2005presence; @makowski2017being]. Indeed,
participants' self-reported emotional arousal were found to
significantly predict the probability that they would perceive images as
real [@azevedo2020body]. Conversely, beliefs that emotional stimuli were
fake (e.g., that emotional scenes were not authentic but instead
involved actors and movie makeup) were found to result in emotion
down-regulation [@makowski2019phenomenal; @sperduti2017distinctive]. In
line with these findings, studies on susceptibility to fake news have
also found heightened stimulus emotionality to be associated with
greater belief [@martel2020reliance; @bago2022emotion], and higher
neurophysiological arousal was predictive of judging realistic images as
real [@azevedo2020body]. Additionally, other factors, such as the
stimuli's perceived self-relevance [@sperduti2016paradox;
@goldstein2009pleasure], as well as familiarity [@begg1992dissociation],
could also play a role in guiding our appraisal of a stimulus. **For
instance,** @miller2023 **reported that participants were more likely to
mistakenly identify AI-generated faces as real because they perceived
them as more familiar.**

@@

Due to their popularity as a target of CGI technology and the prospect
offered with facial features that can be experimentally manipulated,
AI-generated images of faces are increasingly used to study face
processing [@dawel2021systematic], in particular in relationship with
saliency or emotions, as well as to other important components of face
evaluation, such as trustworthiness or attractiveness
[@sobieraj2014beautiful; @balas2017; @tsikandilakis2019beauty;
@calbi2017]. Interestingly, artificially created faces rated as more
attractive (by an independent group of raters) were perceived as less
real [@tucciarelli2020]. Conversely, @liefooghe2022faces reports that
attractiveness ratings were significantly lower when participants who
were told that the faces were AI-generated were compared to those who
had no prior knowledge. **Similarly, when participants are informed that
faces are AI-generated, the perceived artificiality leads to lower trust
ratings [@wang2024], even when they are real faces
[@liefooghe2022faces]. In contrast, when participants are unaware that
the faces are AI-generated, trust ratings for these synthetic faces tend
to increase [@nightingale2022]**. Whereas this line of evidence suggests
that reality beliefs have an effect on face attractiveness and
**trustworthiness** ratings, the opposite question - whether
attractiveness **and trustworthiness contribute** to the formation of
reality beliefs - has received little attention to date.

@@

AI-generated content, in particular realistic images, is becoming
commonplace and carries important risks for misinformation and
black-mailing [@viola2023designed], emphasizing the need to understand
the different components that come into play in the formation of reality
beliefs. This exploratory study primarily aims at investigating the
effect of facial attractiveness on simulation monitoring, i.e., on the
beliefs that an image is real or artificially generated. **The affective
reality theory [@makowski2023affective; @makowski2018cognitive] posits
that the default tendency is to believe that experiences are real, with
emotional and bodily reactions playing a pivotal role in reinforcing or
challenging this belief. According to the theory, there is a quadratic
(inverse U-shaped) relationship between affect and reality judgments:
stimuli that elicit mild to moderate emotional and/or bodily reactions
tend to enhance the perception of realness, increasing confidence in
reality. However, when emotional or bodily responses become too intense
or overwhelming, the default belief shifts towards non-reality beliefs
as an emotion regulation mechanism. In other words, extreme emotional
reactions can potentially trigger beliefs of reality denial (“it cannot
be real”) as a protective mechanism to help individuals cope with
distress. The present experiment can be put in relation with the first
part of this proposal - that of a positive relationship between embodied
or emotional reactions and appraisals of reality. Following this
hypothesis, faces rated as either highly attractive or unattractive -
and eliciting stronger reactions - would likely be judged as real**.
**We expect a similar pattern with trustworthiness, where faces judged
as highly trustworthy or untrustworthy will be more likely to be
perceived as real. Finally, we anticipate a positive relationship
between familiarity and perceived realness, as familiar faces tend to be
judged as more real [@miller2023].** Additionally, we will further
explore the link shared by dispositional traits, such as personality and
attitude towards AI, with simulation monitoring tendencies. Importantly,
this study does investigate the discriminative accuracy between "true"
photos and "true" artificially-generated images (which we consider more
a technological issue than a psychological one), focusing on the beliefs
that a stimulus is real or fake, independently of its true. **In other
words, the present study investigates the psychological process that
leads to different *beliefs* of reality, rather than the discrimination
between real faces and actual AI-generated ones, which largely depends
on the technological quality of the AI-generation process.**

@@

**Procedure.** In the first part of the study, participants answered a
series of personality questionnaires presented in the order below. These
include the *Mini-IPIP6* <!-- (24 items) [@sibley2011]  --> [24 items,
@sibley2011] measuring 6 personality traits, the *SIAS-6* and the
*SPS-6* <!-- (6 items each) [@peters2012development] --> [6 items each,
@peters2012development] assessing social anxiety levels, 5 items we
devised pertaining to expectations about AI-generated image technology
("I think current Artificial Intelligence algorithms can generate very
realistic images"), **to potentially test and mitigate the potential
effect of expectations/beliefs about AI.** **These items were** mixed
with 5 items from the general attitudes towards AI scale to lower the
former's saliency and the possibility of it priming the subjects about
the task, <!-- (*GAAIS*) [@schepman2020initial]. --> [*GAAIS*,
@schepman2020initial] the *FFNI-BF*
<!-- (30 items) [@jauk2022validation] --> [30 items,
@jauk2022validation] measuring 9 facets of narcissism; the *R-GPTS*
<!-- (18 items) [@freeman2021revised] --> [18 items,
@freeman2021revised] measuring 2 dimensions related to paranoid
thinking; and the *IUS-12* <!-- (12 items) [@carleton2007fearing] -->
[12 items, @carleton2007fearing] measuring intolerance to uncertainty.
Self-rated attractiveness was also assessed using 2 items - one
measuring general attractiveness
<!-- ("How attractive would you say you are?") [@marcinkowska2021self] -->["How
attractive would you say you are?", @marcinkowska2021self] and the other
measuring physical
attractiveness<!-- ("How would you rate your own physical attractiveness relative to the average") [@spielmann2020predictive] -->
["How would you rate your own physical attractiveness relative to the
average", @spielmann2020predictive]. 3 attention check questions were
also embedded in the surveys. **All Cronbach's alpha values were within
the acceptable to excellent range, except for the neuroticism subscale
of the Mini-IPIP6 and the negative subscale of the GAAI, which were
poor, and the Expectations about AI scale, which was questionable (Gliem
& Gliem, 2003; see supplementary material for the details of the
reliability analysis).**

In the second part of this study, images of neutral-expression faces
from the validated American Multiracial Face Database
<!-- (AMFD) [@chen2021broadening]  -->[AMFD, @chen2021broadening] were
presented to the participants for 500ms each, in a randomized order,
following a fixation cross display (750 ms). T**he decision to present
the faces for 500 ms was based on pilot studies, which demonstrated that
this duration provides a sufficient perceptual window for
decision-making and aligns with previous research indicating stable
judgment levels and increased confidence beyond this exposure time
[@willis2006first].**

**The AMFD is a recently validated database including a set of 110
pictures of homogeneous quality featuring diverse faces (particularly in
terms of ethnicity), each (except one) posing with either a neutral or
smiling expression. We selected all 109 neutral images (89 women and 20
men) to reduce the influence of confounding factors like affect. The
AMFD primarily features racially ambiguous faces, representing multiple
racial categories such as multiracial, Latinx, and white. The database
includes 81 faces from individuals self-reporting two racial backgrounds
and 29 from those with three or more racial backgrounds: 33%
Asian/White, 22% Latinx/White, 11% Asian/Latinx, 6% White/Middle
Eastern, 5% Black/White, and 5% Asian/Middle Eastern, with about 18%
identifying as other racial backgrounds.**

After each stimulus presentation, ratings of *Trustworthiness* ("I find
this person trustworthy") and *Familiarity* ("This person reminds me of
someone I know") were collected using visual analog scales. Notably, as
facial attractiveness is a multidimensional construct, encompassing
evolutionary, sociocultural, biological as well as cognitive aspects
[@rhodes2006evolutionary; @han2018beauty], we assessed attractiveness
using 2 visual analog scales, measuring general *Attractiveness* ("I
find this person attractive") and physical *Beauty* ("This face is
good-looking"). **This dual-scale approach aims to reflect two
conceptually distinct dimensions: Attractiveness might capture personal,
Self-relevant and subjective appeal, whereas Beauty might be related to
a more “objective” decision based on aesthetic criteria that can be
recognized independently of personal attraction. In other words, we
wanted the experiment to be able to potentially capture scenarios where
a face could be judged beautiful yet not, attractive and vice versa.**

@@

**Participants.** Although the main part of the study relied on
within-subject design (with 109 trials per participant), we also planned
to do between-participants analyses, thus aiming at collecting a larger
sample than traditionally used in experimental psychology (with budget
availability as the main constraint). One hundred and fifty participants
were recruited via *Prolific*, a crowd-sourcing platform recognized for
providing high quality data [@peer2022; douglas2023data]. The only
inclusion criterion was a fluent proficiency in English to ensure that
the experiment instructions would be well-understood. Participants were
incentivised with a reward of about £ 7.5 for completing the study,
which took about 45 minutes to finish. Demographic variables (age,
gender, sexual orientation, education and ethnicity) were self-reported
on a voluntar

We excluded 5 participants that either failed 2 (\>= 66.6%) or more
attention check questions, took an implausibly short time to finish the
questionnaires or had incomplete responses. **Out of the 5 participants
excluded, 2 participants were excluded because they failed 2 out of 3
attention checks, 1 because they did not answer the sexual orientation
question, which made further analysis impossible, and 2 had an abnormal
low agreement (r \< 0.1) between the beauty and the attractiveness
ratings (possibly indicating random responses as these two scales
exhibited a higher correlation for the other participants)**.The final
sample included 145 participants (Mean age = 28.3, SD = 9.0, range:
\[19, 66\]; Sex:48.3% females, 51.0% males, 0.7% others).

**Data Analysis.** The real-fake ratings (measured originally with a
\[-1, 1\] analog scale) were converted into two scores, corresponding to
two conceptually distinct mechanisms: the dichotomous *belief* (real or
fake, based on the sign of the rating) and the *confidence* (the
rating's absolute value) associated with that belief. The former was
analyzed using logistic mixed models, which modelled the probability of
assigning a face to the real (\>= 0) as opposed to fake (\< 0). The
latter, as well as the other face ratings (attractiveness, beauty,
trustworthiness and familiarity), was modelled using mixed beta
regressions (suited for outcome variables expressed in percentages). The
models included the participants and stimuli as random **intercepts with
no nested variables**.

@@

The determinants of reality beliefs were modelled separately for
attractiveness, beauty, trustworthiness, and familiarity, using second
order raw polynomials coefficients to allow for possible quadratic
relationships (**Figure 2**). Aside from attractiveness (conceptualized
as a general construct), models for beauty, trustworthiness and
familiarity were adjusted for the two remaining variables *mutatis
mutandis*. The analysis focused on sexual-orientation relevant stimuli,
i.e., on faces that were aligned with respect to the participants'
sexual orientation (i.e., female faces for heterosexual males, male
faces for homosexual males, etc.), and the models included the
interaction with the participants' gender (as a sexual dimorphism has
been reported in face appraisal processes). For the attractiveness and
beauty models, we then added the interaction with the reported
self-attractiveness (the average of the two questions pertaining to it)
to investigate its potential modulatory effect. Finally, we investigated
the inter-individual correlates of simulation monitoring with similar
models (but this time, for all items regardless of the participant's
gender or sexual orientation) for each questionnaire, with all of the
subscales as orthogonal predictors.\

```{r message=FALSE, warning=FALSE, fig.cap="Figure 2. Top part shows the effect of face ratings on 1) the probability of judging a face as real vs. fake (solid line) and 2) on the confidence associated with that judgement (dashed lines) depending on the sex. Bottom part shows the effect of personality traits on the belief (black line) and the confidence associated with it (colored lines). The points are the average per participant confidence for both types of judgements. Stars indicate significance (p < .001***, p < .01**, p < .05*).", out.width="100%"}
knitr::include_graphics("../../figures/Figure2.png")
```

@@

Several limitations have to be noted. The current experimental paradigm
required participants to judge the realness of faces they had prior
exposure to (which was done to prevent reality judgements from
influencing the other ratings). Although the effect of re-exposure delay
was negligible, the potential bias induced by face familiarity, **that
is by re-presenting the same face stimuli twice**, as compared to
judging completely new items, cannot be discarded. Future studies could
examine that by incorporating novel face images or increasing the
duration of the re-exposure delay.

Another issue is the impact on reality judgements of the prior explicit
instruction that "about half of the faces were AI-generated and the
other half real photos". Given this prior information given to
participants, it might seem like our enthusiasm pertaining to the
finding that most people did indeed believe a high number of stimuli to
be fake might be unwarranted, since it simply affirms participants
followed the instructions. However, even if that was the case, the
finding that our beliefs of reality can be so easily re-programmed with
simple instructions and lead to high-confidence answers remains an
interesting phenomenon. **Moreover, it is to note that the paradigm did
not explicitly instruct participants to balance their answers according
to a certain distribution (e.g., 50-50) - merely providing them a
description of the dataset (but participants could, and in some cases
did, deviate substantially from the information provided).** The fact
that no presentation order effect was found on reality beliefs suggests
that participants did not try to actively distribute their responses to
match the instructions, in which case we would have expected a different
pattern: for instance, the first few items judged as real (the initial
"true" belief of the participants), and a bias would progressively
appear towards responding "fake" (as participants realize that all
stimuli are of similar nature and that they have to "make up" for the
prevalence of their "real" answers to fulfil the expected proportion of
responses given the instructions).

@@

Finally, it is important to note that although consistent in their
directions across models and variables, the magnitude of the effects
found in the study was relatively small, suggesting that the facial
appraisals measured in the study were not the key determinants of
simulation monitoring. Hence, beyond exploring new potential mechanisms,
future studies should include a more thorough debriefing to try to
capture what conscious strategies (if any) the participants used (e.g.,
focusing on some features of the stimulus - like hair or eyes in the
case of faces) to guide their reality beliefs. **Additionally, the role
of specific facial features, like perceived dominance, warmth or gender,
would be an interesting avenue to explore in future studies, in
particular with paradigms directly manipulating these dimensions (for
instance using AI to generate faces of different characteristics).**

>>>>

Advancements in technology have now made it possible to create
near-perfect simulations that are indistinguishable from reality with an
ease, affordability and accessibility that are unprecedented in Human
history. These artificial, yet realistic constructs permeate all areas
of life through immersive works of fiction, deep fakes (real-like images
and videos generated by deep learning algorithms), virtual and augmented
reality (VR and AR), artificial beings (artificial intelligence "bots"
with or without a physical form), fake news and skewed narratives, of
which ground truth is often hard to access [@nightingale2022]. Such
developments not only carry important consequences for the technological
and entertainment sectors, but also for security and politics - for
instance if used for propaganda and disinformation, recruitment into
malevolent organizations, or religious indoctrination [@pantserev2020].
This issue is central to what has been coined the "post-truth era"
[@lewandowsky2017beyond], in which the distinction (and lack thereof)
between authentic and simulated objects will play a critical role.

@@

While not all simulations have achieved perfect realism, such as
Computer Generated Images (CGI) in movies or via recent algorithms such
as GANs or diffusion model, which often include distortions or lack
certain key details distinguishing them from real images
[@mcdonnell2010face; @corvi2022detection], it is fair to assume that
these technical limitations will become negligible in the near future.
This is particularly true in the field of face generation, where
face-generation algorithms are already able to create stimuli that are
virtually indistinguishable from real photos [@nightingale2022;
@tucciarelli2020; @moshel2022]. Such a technological feat, however,
leads to a new question: if real and fake stimuli cannot be
differentiated based on their objective "physical" characteristics, how
can we form judgements regarding their nature?

@@

Literature shows that the context surrounding a stimulus often plays an
important role in the assessment of its reality [a process henceforth
referred to as simulation monitoring, @makowski2018cognitive;
@makowski2019phenomenal]. With the extensive search and processing of
cues within ambiguous stimuli being an increasingly complex and
cognitively effortful strategy [@michael2021source;
@susmann2021persuasion], people tend to draw on peripheral contextual
cues (**Figure 1**), such as the source of the stimulus (e.g., which
journal was the information published in), and its credibility,
authority and expertise, to help facilitate their evaluation
[@petty1986elaboration; @susmann2021persuasion; @michael2021source].
However, the atomization and decontextualization of information allowed
by online social media (where text snippets or video excerpts are often
mass-shared with little context) makes this task progressively difficult
[@berghel2018weaponizing; @chen2015news]. Thus, in the absence of clear
contextual information, what drives our beliefs of reality?

@@

Evidence suggests that inter-individual characteristics play a crucial
role in simulation monitoring, with factors such as cognitive style,
prior beliefs, and personality traits [@bryanov2021determinants;
@ecker2022psychological; @sindermann2020short]. For instance,
individuals with stronger analytical reasoning skills have been found to
better discriminate real from fake stimuli [@pehlivanoglu2021role;
@pennycook2019lazy], and prior knowledge or beliefs about the stimulus
influences one's perception of it by biasing the attention deployment
towards information that is in line with one's expectations
[@britt2019reasoned]. Furthermore, dispositional traits, such as high
levels of narcissism and low levels of openness and conscientiousness,
have been associated with greater susceptibility to fake news
[@piksa2022cognitive; @sindermann2020short]. Interestingly, a recent
review suggested that narcissism was related to a strong self-perceived
ability at detecting lies [@turi2022tangled], which could translate to
participants scoring high on narcissism providing more clear cut and
confidence responses. Conversely, those high in honesty-humility tend to
be more conservative in their judgments to ensure fairness
[@liu2020honesty], likely resulting in lower confidence ratings.

@@

Beyond stimulus- and individual-related characteristics, evidence
suggests that the interaction between the two (i.e., the subjective
reaction associated with the experience of a given stimulus),
contributes to simulation monitoring decisions. For instance, the
intensity of experienced emotions have been shown to increase one's
sense of presence - the extent to which one feels like "being there", as
if the object of experience was real - when engaged in a fictional movie
or a VR environment [@sanchez2005presence; @makowski2017being]. Indeed,
participants' self-reported emotional arousal were found to
significantly predict the probability that they would perceive images as
real [@azevedo2020body]. Conversely, beliefs that emotional stimuli were
fake (e.g., that emotional scenes were not authentic but instead
involved actors and movie makeup) were found to result in emotion
down-regulation [@makowski2019phenomenal; @sperduti2017distinctive]. In
line with these findings, studies on susceptibility to fake news have
also found heightened stimulus emotionality to be associated with
greater belief [@martel2020reliance; @bago2022emotion], and higher
neurophysiological arousal was predictive of judging realistic images as
real [@azevedo2020body]. Additionally, other factors, such as the
stimuli's perceived self-relevance [@sperduti2016paradox;
@goldstein2009pleasure], as well as familiarity [@begg1992dissociation],
could also play a role in guiding our appraisal of a stimulus. For
instance, @miller2023 reported that participants were more likely to
mistakenly identify AI-generated faces as real because they perceived
them as more familiar.

@@

Due to their popularity as a target of CGI technology and the prospect
offered with facial features that can be experimentally manipulated,
AI-generated images of faces are increasingly used to study face
processing [@dawel2021systematic], in particular in relationship with
saliency or emotions, as well as to other important components of face
evaluation, such as trustworthiness or attractiveness
[@sobieraj2014beautiful; @balas2017; @tsikandilakis2019beauty;
@calbi2017]. Interestingly, artificially created faces rated as more
attractive (by an independent group of raters) were perceived as less
real [@tucciarelli2020]. Conversely, @liefooghe2022faces reports that
attractiveness ratings were significantly lower when participants who
were told that the faces were AI-generated were compared to those who
had no prior knowledge. Similarly, when participants are informed that
faces are AI-generated, the perceived artificiality leads to lower trust
ratings [@wang2024], even when they are real faces
[@liefooghe2022faces]. In contrast, when participants are unaware that
the faces are AI-generated, trust ratings for these synthetic faces tend
to increase [@nightingale2022]. Whereas this line of evidence suggests
that reality beliefs have an effect on face attractiveness and
trustworthiness ratings, the opposite question - whether
attractiveness and trustworthiness contribute to the formation of
reality beliefs - has received little attention to date.

@@

AI-generated content, in particular realistic images, is becoming
commonplace and carries important risks for misinformation and
black-mailing [@viola2023designed], emphasizing the need to understand
the different components that come into play in the formation of reality
beliefs. This exploratory study primarily aims at investigating the
effect of facial attractiveness on simulation monitoring, i.e., on the
beliefs that an image is real or artificially generated. The affective
reality theory [@makowski2023affective; @makowski2018cognitive] posits
that the default tendency is to believe that experiences are real, with
emotional and bodily reactions playing a pivotal role in reinforcing or
challenging this belief. According to the theory, there is a quadratic
(inverse U-shaped) relationship between affect and reality judgments:
stimuli that elicit mild to moderate emotional and/or bodily reactions
tend to enhance the perception of realness, increasing confidence in
reality. However, when emotional or bodily responses become too intense
or overwhelming, the default belief shifts towards non-reality beliefs
as an emotion regulation mechanism. In other words, extreme emotional
reactions can potentially trigger beliefs of reality denial (“it cannot
be real”) as a protective mechanism to help individuals cope with
distress. The present experiment can be put in relation with the first
part of this proposal - that of a positive relationship between embodied
or emotional reactions and appraisals of reality. Following this
hypothesis, faces rated as either highly attractive or unattractive -
and eliciting stronger reactions - would likely be judged as real.
We expect a similar pattern with trustworthiness, where faces judged
as highly trustworthy or untrustworthy will be more likely to be
perceived as real. Finally, we anticipate a positive relationship
between familiarity and perceived realness, as familiar faces tend to be
judged as more real [@miller2023]. Additionally, we will further
explore the link shared by dispositional traits, such as personality and
attitude towards AI, with simulation monitoring tendencies. Importantly,
this study does investigate the discriminative accuracy between "true"
photos and "true" artificially-generated images (which we consider more
a technological issue than a psychological one), focusing on the beliefs
that a stimulus is real or fake, independently of its true. In other
words, the present study investigates the psychological process that
leads to different beliefs of reality, rather than the discrimination
between real faces and actual AI-generated ones, which largely depends
on the technological quality of the AI-generation process.

@@

**Procedure.** In the first part of the study, participants answered a
series of personality questionnaires presented in the order below. These
include the *Mini-IPIP6* <!-- (24 items) [@sibley2011]  --> [24 items,
@sibley2011] measuring 6 personality traits, the *SIAS-6* and the
*SPS-6* <!-- (6 items each) [@peters2012development] --> [6 items each,
@peters2012development] assessing social anxiety levels, 5 items we
devised pertaining to expectations about AI-generated image technology
("I think current Artificial Intelligence algorithms can generate very
realistic images"), to potentially test and mitigate the potential
effect of expectations/beliefs about AI.These items were mixed
with 5 items from the general attitudes towards AI scale to lower the
former's saliency and the possibility of it priming the subjects about
the task, <!-- (*GAAIS*) [@schepman2020initial]. --> [*GAAIS*,
@schepman2020initial] the *FFNI-BF*
<!-- (30 items) [@jauk2022validation] --> [30 items,
@jauk2022validation] measuring 9 facets of narcissism; the *R-GPTS*
<!-- (18 items) [@freeman2021revised] --> [18 items,
@freeman2021revised] measuring 2 dimensions related to paranoid
thinking; and the *IUS-12* <!-- (12 items) [@carleton2007fearing] -->
[12 items, @carleton2007fearing] measuring intolerance to uncertainty.
Self-rated attractiveness was also assessed using 2 items - one
measuring general attractiveness
<!-- ("How attractive would you say you are?") [@marcinkowska2021self] -->["How
attractive would you say you are?", @marcinkowska2021self] and the other
measuring physical
attractiveness<!-- ("How would you rate your own physical attractiveness relative to the average") [@spielmann2020predictive] -->
["How would you rate your own physical attractiveness relative to the
average", @spielmann2020predictive]. 3 attention check questions were
also embedded in the surveys. All Cronbach's alpha values were within
the acceptable to excellent range, except for the neuroticism subscale
of the Mini-IPIP6 and the negative subscale of the GAAI, which were
poor, and the Expectations about AI scale, which was questionable (Gliem
& Gliem, 2003; see supplementary material for the details of the
reliability analysis).

In the second part of this study, images of neutral-expression faces
from the validated American Multiracial Face Database
<!-- (AMFD) [@chen2021broadening]  -->[AMFD, @chen2021broadening] were
presented to the participants for 500ms each, in a randomized order,
following a fixation cross display (750 ms). The decision to present
the faces for 500 ms was based on pilot studies, which demonstrated that
this duration provides a sufficient perceptual window for
decision-making and aligns with previous research indicating stable
judgment levels and increased confidence beyond this exposure time
[@willis2006first].

The AMFD is a recently validated database including a set of 110
pictures of homogeneous quality featuring diverse faces (particularly in
terms of ethnicity), each (except one) posing with either a neutral or
smiling expression. We selected all 109 neutral images (89 women and 20
men) to reduce the influence of confounding factors like affect. The
AMFD primarily features racially ambiguous faces, representing multiple
racial categories such as multiracial, Latinx, and white. The database
includes 81 faces from individuals self-reporting two racial backgrounds
and 29 from those with three or more racial backgrounds: 33%
Asian/White, 22% Latinx/White, 11% Asian/Latinx, 6% White/Middle
Eastern, 5% Black/White, and 5% Asian/Middle Eastern, with about 18%
identifying as other racial backgrounds.

After each stimulus presentation, ratings of *Trustworthiness* ("I find
this person trustworthy") and *Familiarity* ("This person reminds me of
someone I know") were collected using visual analog scales. Notably, as
facial attractiveness is a multidimensional construct, encompassing
evolutionary, sociocultural, biological as well as cognitive aspects
[@rhodes2006evolutionary; @han2018beauty], we assessed attractiveness
using 2 visual analog scales, measuring general *Attractiveness* ("I
find this person attractive") and physical *Beauty* ("This face is
good-looking"). This dual-scale approach aims to reflect two
conceptually distinct dimensions: Attractiveness might capture personal,
Self-relevant and subjective appeal, whereas Beauty might be related to
a more “objective” decision based on aesthetic criteria that can be
recognized independently of personal attraction. In other words, we
wanted the experiment to be able to potentially capture scenarios where
a face could be judged beautiful yet not, attractive and vice versa.

@@

**Participants.** Although the main part of the study relied on
within-subject design (with 109 trials per participant), we also planned
to do between-participants analyses, thus aiming at collecting a larger
sample than traditionally used in experimental psychology (with budget
availability as the main constraint). One hundred and fifty participants
were recruited via *Prolific*, a crowd-sourcing platform recognized for
providing high quality data [@peer2022; douglas2023data]. The only
inclusion criterion was a fluent proficiency in English to ensure that
the experiment instructions would be well-understood. Participants were
incentivised with a reward of about £ 7.5 for completing the study,
which took about 45 minutes to finish. Demographic variables (age,
gender, sexual orientation, education and ethnicity) were self-reported
on a voluntary basis.

We excluded 5 participants that either failed 2 (\>= 66.6%) or more
attention check questions, took an implausibly short time to finish the
questionnaires or had incomplete responses. Out of the 5 participants
excluded, 2 participants were excluded because they failed 2 out of 3
attention checks, 1 because they did not answer the sexual orientation
question, which made further analysis impossible, and 2 had an abnormal
low agreement (r \< 0.1) between the beauty and the attractiveness
ratings (possibly indicating random responses as these two scales
exhibited a higher correlation for the other participants).The final
sample included 145 participants (Mean age = 28.3, SD = 9.0, range:
\[19, 66\]; Sex:48.3% females, 51.0% males, 0.7% others).

**Data Analysis.** The real-fake ratings (measured originally with a
\[-1, 1\] analog scale) were converted into two scores, corresponding to
two conceptually distinct mechanisms: the dichotomous *belief* (real or
fake, based on the sign of the rating) and the *confidence* (the
rating's absolute value) associated with that belief. The former was
analyzed using logistic mixed models, which modelled the probability of
assigning a face to the real (\>= 0) as opposed to fake (\< 0). The
latter, as well as the other face ratings (attractiveness, beauty,
trustworthiness and familiarity), was modelled using mixed beta
regressions (suited for outcome variables expressed in percentages). The
models included the participants and stimuli as random **intercepts with
no nested variables.

@@

The determinants of reality beliefs were modelled separately for
attractiveness, beauty, trustworthiness, and familiarity, using second
order raw polynomials coefficients to allow for possible quadratic
relationships (**Figure 2**). Aside from attractiveness (conceptualized
as a general construct), models for beauty, trustworthiness and
familiarity were adjusted for the two remaining variables *mutatis mutandis*. The analysis focused on sexual-orientation relevant stimuli,
i.e., on faces that were aligned with respect to the participants'
sexual orientation (i.e., female faces for heterosexual males, male
faces for homosexual males, etc.), and the models included the
interaction with the participants' gender (as a sexual dimorphism has
been reported in face appraisal processes). For the attractiveness and
beauty models, we then added the interaction with the reported
self-attractiveness (the average of the two questions pertaining to it)
to investigate its potential modulatory effect. Finally, we investigated
the inter-individual correlates of simulation monitoring with similar
models (but this time, for all items regardless of the participant's
gender or sexual orientation) for each questionnaire, with all of the
subscales as orthogonal predictors.

```{r message=FALSE, warning=FALSE, fig.cap="Figure 2. Top part shows the effect of face ratings on 1) the probability of judging a face as real vs. fake (solid line) and 2) on the confidence associated with that judgement (dashed lines) depending on the sex. Bottom part shows the effect of personality traits on the belief (black line) and the confidence associated with it (colored lines). The points are the average per participant confidence for both types of judgements. Stars indicate significance (p < .001\\*\\*\\*, p < .01\\*\\*, p < .05\\*).", out.width="100%"}
knitr::include_graphics("../../figures/Figure2.png")
```

@@

Several limitations have to be noted. The current experimental paradigm
required participants to judge the realness of faces they had prior
exposure to (which was done to prevent reality judgements from
influencing the other ratings). Although the effect of re-exposure delay
was negligible, the potential bias induced by face familiarity, that
is by re-presenting the same face stimuli twice, as compared to
judging completely new items, cannot be discarded. Future studies could
examine that by incorporating novel face images or increasing the
duration of the re-exposure delay.

Another issue is the impact on reality judgements of the prior explicit
instruction that "about half of the faces were AI-generated and the
other half real photos". Given this prior information given to
participants, it might seem like our enthusiasm pertaining to the
finding that most people did indeed believe a high number of stimuli to
be fake might be unwarranted, since it simply affirms participants
followed the instructions. However, even if that was the case, the
finding that our beliefs of reality can be so easily re-programmed with
simple instructions and lead to high-confidence answers remains an
interesting phenomenon. Moreover, it is to note that the paradigm did
not explicitly instruct participants to balance their answers according
to a certain distribution (e.g., 50-50) - merely providing them a
description of the dataset (but participants could, and in some cases
did, deviate substantially from the information provided). The fact
that no presentation order effect was found on reality beliefs suggests
that participants did not try to actively distribute their responses to
match the instructions, in which case we would have expected a different
pattern: for instance, the first few items judged as real (the initial
"true" belief of the participants), and a bias would progressively
appear towards responding "fake" (as participants realize that all
stimuli are of similar nature and that they have to "make up" for the
prevalence of their "real" answers to fulfill the expected proportion of
responses given the instructions).

@@

Finally, it is important to note that although consistent in their
directions across models and variables, the magnitude of the effects
found in the study was relatively small, suggesting that the facial
appraisals measured in the study were not the key determinants of
simulation monitoring. Hence, beyond exploring new potential mechanisms,
future studies should include a more thorough debriefing to try to
capture what conscious strategies (if any) the participants used (e.g.,
focusing on some features of the stimulus - like hair or eyes in the
case of faces) to guide their reality beliefs. Additionally, the role
of specific facial features, like perceived dominance, warmth or gender,
would be an interesting avenue to explore in future studies, in
particular with paradigms directly manipulating these dimensions (for
instance using AI to generate faces of different characteristics).
